simplex: An individual-based ecological modeling platform

### 2015-10-26 paper/PrePrint.Rmd
modified PrePrint.Rmd and generated new PrePrint.pdf -KL

---
title: "simplex: An individual-based modeling platform"
author: "Kenneth J. Locey,
Dept. of Biology, University of Indiana, Bloomington"
geometry: margin=2.54cm
output: pdf_document
header-includes:
    - \usepackage{setspace}
    - \doublespacing
---

@@

Here, we present source code and a detailed description for an ecological IBM platform (called simplex) that simulates a broad range of ecological conditions, records information from individuals to the ecosystem, and provides computing code for analyzing output. This platform incorporates computational fluid dynamics (Succi 2001), life history processes and concepts and patterns from biogeography (Hubbell 2001, Bell 2001), evolution and aspects of selection (Hartl and Clark 1997), and integrates nutrient-limited growth and physiology (Pirt 1965, Droop 1983) among dimension of ecology and evolutionary science. 
We constructed simplex with the aim of simulating ecologically complex scenarios, from which the results could be used to test (*i*) predictions of ecological theory and (*ii*) the general influence of constraints and processes. 
Below, we provide detailed explanation of how hydrobide works, the data it quantifies and tracks, the theories and principles hydrobide integrates, and the analyses that can be conducted using the code we provide.

@@

**Overview**--simplex was developed to accomplish three objectives. First, simplex assembles and runs individual-based ecological models from random combinations of state variables and processes. Second, simplex stores model output as R-style data frames and, when the user chooses, generates and stores animations of models. Third, simplex provides R-based files for analyzing simulated; these will eventually be offered in Python as well.

@@

**Theories**
*Constraint-based theory*: simplex was originally built to explore the influence of ecosystem residence time (volume/flow rate) on community assembly and structure. That is, the idea that both ecological processes and constraints shape ecological diversity.

@@

## Output data
simplex generates six files as its output data. They are:

@@

*IndRTD.csv*, *ResRTD.csv*, *TracerRTD.csv*--Each line of these three files contains the amount of time that each individual, resource particle, or inert tracer (from beginning to end) spent in the system. Each line is a run of a randomly assembled model.

@@

@@

The simplex platform assembles individual-based models from random combinations of state-variables and processes. This is done to explore a large degree of variation in ecological conditions and complexity. In this way, simplex allows ecologists to examine whether the influence of a particular variable should be robust to a wide range of ecological conditions.

By taking advantage of modern computing and version control and social coding tools (e.g., Python, R/RStudio, Knitr, Git, GitHub) simplex is well-positioned to keep developing. Future developments will include additional ecological dynamcis such as predator-prey, mutualism, and parasitism. Improvements to simplex will also include an explicit stoichiometry, where individual resource particles have ratios of Carbon, Nitrogen, and Phosphorus, as well as degrees of biocomplexity. For example, given the chemical structure of phosphate and the size of a resource particle, the particle can be assigned *x* units of phosphorus as well as the biocomplexity value of phosphate (estimated as a form of Shannon's entropy).

Likewise, improvements and future versions of simplex will provide increasing numbers of files for statistical analysis of simplex output. These files will, in a sense, be able to be used as a complete analysis of some particular aspect of simplex output (e.g., demographic, resource-related, spatial, multivariate, ecological model testing). Moreover, these and all simplex files are freely available and open source, allowing users to modify simplex source code to suit their particular needs.

@@

I thank Jay T. Lennon, Mario Muscarella, Xiao Xiao, and Ed Hall for fruitful discussions. I thanks Dr. Daniel V. Schroeder for his web-based fluid dynamic simulator, the python-based code for which serves as the core method in simplex for simulating fluid flow.

>>>>

---
title: "simplex: An individual-based ecological modeling platform"
author: "Kenneth J. Locey,
Dept. of Biology, University of Indiana, Bloomington"
geometry: margin=2.54cm
output: pdf_document
header-includes:
    - \usepackage{setspace}
    - \doublespacing
---

@@

Here, I present source code and a detailed description for an ecological IBM platform (called simplex) that simulates a broad range of ecological conditions, records information from individuals to the ecosystem, and provides computing code for analyzing output. This platform incorporates computational fluid dynamics (Succi 2001), life history processes and concepts and patterns from biogeography (Hubbell 2001, Bell 2001), evolution and aspects of selection (Hartl and Clark 1997), and integrates nutrient-limited growth and physiology (Pirt 1965, Droop 1983) among dimension of ecology and evolutionary science. 
I constructed simplex with the aim of simulating ecologically complex scenarios, from which the results could be used to test (*i*) predictions of ecological theory and (*ii*) the general influence of constraints and processes. 
Below, I provide detailed explanation of how hydrobide works, the data it quantifies and tracks, the theories and principles hydrobide integrates, and the analyses that can be conducted using the code we provide.

@@

**Overview**--I developed simplex to accomplish three objectives. First, simplex assembles and runs individual-based ecological models from random combinations of state variables and processes. Second, simplex stores the output of models as R-style data frames and .csv files, and when the user chooses, generates and stores animations of models. Third, simplex provides R Markdown files for analyzing simulated data; these will eventually be offered in Python as well.

@@

**Theories**
*Constraint-based theory*: simplex was originally built to explore the influence of ecosystem residence time (volume/flow rate) on community assembly and structure. That is, the idea that both ecological processes and physical constraints shape ecological diversity.

@@

## Products
### Output data
simplex generates six files as its output data. They are:

@@

*IndRTD.csv*, *ResRTD.csv*, *TracerRTD.csv*--Each line of these three files contains the amount of time that each individual, resource particle, or inert tracer spent in the system. Each line represents the same run of a randomly assembled model.

@@

### R analysis files

There are currently two R Markdown analysis files associated with simplex; these are located in the `tools/analyses/Rmd` directory. 
These files are primarily designed to help users understand how to import simplex files, to use R Markdown to analyze simplex data, and conduct descriptive and univariate statistics on simplex output. The first file, `BasicAnalysis.Rmd`, focuses on descriptive statistics, multiple univariate analyses, and variance partitioning to reveal relationships among simplex variables. The second file, `Diversity.Rmd`, focuses on the analysis of taxonomic diversity and abundance to include fitting species abundance models using the R-based `vegan` module. Each file generates a knitted pdf (includes code, text, and figures), see Supplement.

### Animations

I used the animation module in the matplotlib plotting library to generate avi formatted movies of simplex models. These can be viewed in any internet brower (https://github.com/LennonLab/simplex/tree/master/results/movies/examples). Currently, generating movies requires following three instructions in model/model.py:

1.) uncomment line 'ani.save' on or near line 364  
2.) adjust the frames variable on or near line 364, to change movie length   
3.) change plot_system = 'no' to 'yes' on or near line 66

Because generating animations requires computing time and memory, doing so can be computationally demanding. To quicken the process, comment out plt.show() on or near line 368.

@@

The simplex platform assembles individual-based models from random combinations of state-variables and processes. This is done to explore a large degree of variation in ecological conditions and complexity. In this way, simplex allows ecologists or educators to examine whether the influence of a particular variable should be robust to a wide range of ecological conditions. In addition to its unique  modeling aspects, simplex includes analysis code that shows the user how to import and analyze the simulated data, and provides a basic demonstration of how to conduct variance partitioning and fitting of species abundance models.

By taking advantage of modern computing and version control and social coding tools (e.g., Python, R/RStudio, Knitr, Git, GitHub) simplex is well-positioned to continue development as both a research and education/instructional tool for ecology. Future developments and released versions will include additional ecological dynamcis such as predator-prey, mutualism, and parasitism. Improvements to simplex will also include an explicit stoichiometry, where individual resource particles have ratios of Carbon, Nitrogen, and Phosphorus, as well as degrees of biocomplexity. For example, given the chemical structure of phosphate and the size of a resource particle, the particle can be assigned *x* units of phosphorus as well as the biocomplexity value of phosphate (estimated as a form of Shannon's entropy).

Likewise, improvements and future versions of simplex will provide increasing numbers of files for statistical analysis of simplex output. These files will, in a sense, be able to be used as a complete analysis of some particular aspect of simplex output (e.g., demographic, resource-related, spatial, multivariate, ecological model testing). These and all simplex files are freely available and open source, allowing users to modify simplex source code to suit their particular needs.

@@

I thank Jay T. Lennon, Mario Muscarella, Xiao Xiao, and Ed Hall for fruitful discussions. I thank Dr. Daniel V. Schroeder for his web-based fluid dynamic simulator, the python-based code for which serves as the core method in simplex for simulating fluid flow.
